```{r echo=FALSE}
species.select <- "RNTR"
Version <- '2017-05-05'
```

## `r species.select`

```{r RNTRSummary, echo=FALSE, message=FALSE, warning=FALSE }
# Comput the summary statstics for each species by year
cpue.select <- cpue[ cpue$Species.Code == species.select,]

fish.stat <- plyr::ddply(cpue.select, c("Species.Code","Year"), plyr::summarize,
                   nfish = length(Fork.Length..mm.),
                   fl.mean=round(mean(Fork.Length..mm., na.rm=TRUE)),
                   fl.min= round(min(Fork.Length..mm. , na.rm=TRUE)),
                   fl.max= round(max(Fork.Length..mm. , na.rm=TRUE)))
fish.stat$fl.mean[ is.nan(fish.stat$fl.mean)    ] <- NA
fish.stat$fl.min [ is.infinite(fish.stat$fl.min)] <- NA
fish.stat$fl.max [ is.infinite(fish.stat$fl.max)] <- NA

# Sort by year  
fish.stat <- fish.stat[ order(fish.stat$Year),]

# Get the Table and Figure numbers
RNTR.tab.fishsummary<-getTABNUM()
RNTR.fig.forklength <-getFIGNUM()

```
The  mean fork length and size range for this species on a yearly basis are summarized in 
Table `r RNTR.tab.fishsummary` 
and plotted in 
Figure `r RNTR.fig.forklength`.

```{r RNTR.fish.summary.tab,  echo=FALSE, results='asis'}
# Create the output table
fish.table <- fish.stat
colnames(fish.table) <- c(
       "Species\nCode",
       "Year",
       "n",
       "Mean\nfork\nlength\nmm",
       "Min\nfork\nlength\nmm",
       "Max\nfork\nlength\nmm"
    )
pandoc.table(fish.table[,-1],
             caption=paste("Table ", RNTR.tab.fishsummary,". Summary statistics on fork length for ",
                           species.select,
                           " captured in ",cpue$WatershedName[1], sep=""),
             justify='rrrrr',
             split.cells=c(1,1,1,1,1))
```


```{r RNTR.forklength.fig,  echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5, dpi=300}
# Create a plot of the change in fork length over all of the years

# Melt the data
sumstat.melt <- reshape2::melt(fish.stat,
                        id.var=c("Species.Code","Year"),
                        measure.var=c("fl.mean","fl.min","fl.max"),
                        variable.name="Measure",
                        value.name="Length")
sumstat.melt$Measure <- car::recode(sumstat.melt$Measure,
                            " 'fl.mean'='Mean'; 'fl.max'='Max'; 'fl.min'='Min' " )

sumstatplot <- ggplot2::ggplot(data=sumstat.melt, aes(x=Year, y=Length, color=Measure))+
   ggtitle(paste('Figure ', RNTR.fig.forklength,'. Trends in fork length for ', species.select,
                 ' on ', cpue.select$WatershedName[1],sep=""))+
   theme(plot.title = element_text(size = 8))+  # change text size for title
   geom_point()+
   geom_line()+
   xlab("Year")+ylab("Length (mm)")
sumstatplot
```


```{r RNTR.lengthdis.fignum, echo=FALSE}
# Get the figure number for the length distribution
fig.RNTR.lengthdist<-getFIGNUM()
```

The length distribution over all years is shown in 
Figure `r fig.RNTR.lengthdist`.
Black vertical line indicates estimated length at 50\% maturity (**999999999999** mm Fork Length).
{\Large Not yet shown --How is this known from the data? }

```{r RNTR.lengthdist, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5, dpi=300}
# Create a histogram of the fork length over all years
lengthdist <- ggplot2::ggplot(data=cpue.select, aes(x=Fork.Length..mm., y=..density..))+
   ggtitle(paste("Figure ", fig.RNTR.lengthdist,". Fork length distribution for ", species.select,' on ',
                 cpue.select$WatershedName[1],sep=""))+
   theme(plot.title = element_text(size = 8))+  # change text size for title
   geom_histogram(alpha=0.2)+ 
   geom_density()+
   xlab("Fork length (mm)")+ylab("Density")
lengthdist
```

```{r RNTR.row.trend.fignum, echo=FALSE}
   # get the figure number for the CPUE plot
   fig.RNTR.rawtrend<-getFIGNUM()
```

A plot of the CPUE over time is shown in 
Figure `r fig.RNTR.rawtrend`.

```{r  RNTR.raw.trend, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5, dpi=300}
catch.summary.select <- catch.summary[ catch.summary$Species.Code == species.select,]

raw.trend <- ggplot2::ggplot(data=catch.summary.select, aes(x=Year, y=CPUE.300m))+
   ggtitle(paste('Figure ', fig.RNTR.rawtrend, ". CPUE data for ", species.select, 
                 " at ", catch.summary.select$WatershedName[1], " with FSI categories",sep=""))+
   theme(plot.title = element_text(size = 8))+  # change text size for title
   geom_point()+
   geom_line(aes(group=LocationTTM))+
   geom_hline(data=FSI.threshold[FSI.threshold$Species.Code == species.select,], 
              aes(yintercept=lower), alpha=0.5)+
   geom_text(data=FSI.threshold[FSI.threshold$Species.Code == species.select,], aes(x=-Inf, y=lower),
              label=FSI.threshold[FSI.threshold$Species.Code == species.select,]$FSI.cat,
              hjust="left", vjust="bottom", color="red")+
   ylab("CPUE (fish/300 m)")+
   xlab("Year\nLines join Locations over time")
raw.trend
```


```{r  RNTR.fit.Bayesian, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# Fit the Bayesian trend model. There are three possible cases.
#  (a) There are 3 or more years of data. We can fit a trend line, estimate process error etc
#  (b) There are only 2 years of data. We need to assume that the system is in steady state (i.e. no trend)
#  (c) There is only 1 year of data. So bad, so sad.
nyears <- length(unique(catch.summary.select$Year))

if(nyears >= 3) {
    fsi.res <- BayesianTrendFSI(catch.summary.select, FSI.threshold)
    slope <- paste(sprintf("%5.2f", 100*fsi.res$beta.table["slope"]),"\\%",sep="")
    SD    <- paste(sprintf("%5.2f", 100*fsi.res$beta.table["sd"])   ,"\\%",sep="")
    p.pos <- paste(sprintf("%5.2f", 1-fsi.res$beta.table["p.slope.lt.0"]))
}
if(nyears == 2){
    fsi.res <- BayesianNoTrendFSI(catch.summary.select, FSI.threshold)
    # slope doesn't exits, and interecept are NOT interesting
}
if(nyears == 1){  
    fsi.res <- BayesianSingleYearFSI(catch.summary.select, FSI.threshold)
    # slope doesn't exits, and interecept are NOT interesting
}
```

```{r RNTR.Bayes.output.3years, eval=(nyears>=3), results='asis', echo=FALSE}
cat("The Bayesian analysis on trend found that the median CPUE was changing at ",sep="")
cat( slope,' (SD ', SD, ') per year ',sep="")
cat("and the posterior probability that the slope is positive is ", p.pos, '.',sep="")
```
```{r RNTR.Bayes.output.2years, eval=(nyears==2), results='asis', echo=FALSE}
cat("With only two years of data, a trend line could not be computed. ")
cat("The Bayesian analysis assumed that there was no change in the median response between the ")
cat("two years to estimate process error and the probability of belonging to each FSI category. ")
```
```{r RNTR.Bayes.output.1years, eval=(nyears==1), results='asis', echo=FALSE}
cat("With only one of data, a trend line could not be computed. ")
cat("Nor can year-specific effects (process error) be separated from site-effects. ")
cat("The FSI categorization may be an artefact of this particular year's set of data. Use with caution. ")
```

```{r RNTR.postplot.fignum, echo=FALSE}
   # get the figure numbers for the posterior plot and the fsi plot
   fig.RNTR.postplot<-getFIGNUM()
   fig.RNTR.fsiplot <-getFIGNUM()
```

Plots of the posterior distribution of `r if(nyears>=3)cat(" the trend line for ")`  the median and the FSI Category membership are shown in 
Figure `r fig.RNTR.postplot` and
Figure `r fig.RNTR.fsiplot`.


```{r  RNTR.postplot, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5, dpi=300}
fsi.res$postplot+
    ylim(c(0,150))+
   ggtitle(paste('Figure ', fig.RNTR.postplot, ". Estimated MEDIAN trend line for ", species.select,  
                 " at ", catch.summary.select$WatershedName[1], " with FSI categories",sep=""))+
   theme(plot.title = element_text(size = 8))+  # change text size for title
   ylab("CPUE per 300 m")
```


```{r  RNTR.fsi.plot, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3, fig.width=5, dpi=300}
#caption{`r paste("Posterior probability of membership in FSI category for  ", 
#                 catch.summary.select$Species.Code[1], 
#                " at ", catch.summary.select$WatershedName[1],sep="")`

fsi.res$fsi.plot+
   ggtitle(paste('Figure ', fig.RNTR.fsiplot, ". Posterior probability by FSI category for ", species.select,  
                 " at ", catch.summary.select$WatershedName[1],sep=""))+
   theme(plot.title = element_text(size = 8))+  # change text size for title
   ylab("Cumulative Probability")
```



